Prestige as a Driving Force in Cultural Transmission

### 2022-04-05 ms/PrestigeBias.tex
update

\begin{abstract}
Copying role-models can be an efficient method for acquiring knowledge. A common bias when choosing a role-model to copy is success bias: copying whoever appears more successful. This bias depends on the performance of the role-model alone, with no other factors. We propose an additional bias that may be prevalent in cultural transmission: influence bias, in which role-model choice is affected by the number of individuals that have already copied each potential role-model. We combine success and influence bias together to a ``prestige bias'' and analyze its effects on cultural-evolutionary dynamics using mathematical analysis and stochastic simulations. We find analytic approximations to our stochastic model, facilitating further mathematical analysis and reducing the computational complexity of simulations. We validate these approximations using simulations, and demonstrate their robustness to model assumptions.
We also find approximations to the fixation probability and the fixation time of an invading advantageous cultural trait, in both a constant and changing environment, which resemble Kimura's classical formulas.
These approximations show that success bias effectively plays the part of natural selection, whereas influence bias effectively reduces the population size.	
It also accelerates the evolutionary dynamics, as can be expected in a \textit{rich-getting-richer} process.
Our model may provide a better description of human cultural transmission, 
especially in the last years where social networks are very popular. % TODO to discuss this last sentence
Further work is needed to test if this model could predict various phenomena in human cultural evolution when extended with the effects of selection and innovation.
\end{abstract}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}

@@

\paragraph{Success bias.}
\citet[Ch.8, p.247-249]{evolutionBook} describe a blended transmission algorithm by defining $F$ as a weighted average of the traits of all role-models, 
\begin{equation}\label{eq:boydF}
% TODO I changed X to A, and Aij to Aj, verify its ok
F_i(\vec{A}) = \sum_{j=1}^N G_{i,j}\cdot A_{j} \;, 
\end{equation}
where $G_{i,j}$ is the success bias of role-model $j$ in the eyes of copier $i$,
\begin{equation}\label{eq:boydG}
G_{i,j} = \frac{\beta(A_{i,j})}{\sum_{k=1}^{N} \beta(A_{i,k})} \;,
\end{equation}
$A_{i,j}$ is the absolute indicator trait value copier $i$ estimates for role-model $j$ with some error $e_i \sim N(0,\frac{1}{\eta^2})$,
\begin{equation}\label{eq:relativeIndicator}
A_{i,j} = A_j + e_i,
\end{equation}
and $\beta(\cdot)$ is the bias function that quantifies the success bias of a role-model,
\begin{equation}\label{eq:success_bias}
\beta(A_{i,j}) = b \cdot \exp{\Big(-\frac{(A_{i,j} - \hat{A})^2}{2J}\Big)},
\end{equation} 
with $\hat{A}$ as the arbitrary optimal indicator value, and $J$ and $b$ as parameters that control the bias strength.
Therefore, $G_{i,j}$ is a relative success score that copier $i$ assigns to role-model $j$, equivalent to relative fitness in evolutionary-genetic transmission models.

@@

\subsection*{Dichotomous trait}
To allow for mathematical analysis of the model, we introduce a simplified version where the indicator trait only has two phenotypes: the optimal phenotype $\hat{A}$ and the sub-optimal phenotype $A$. 
All role-models with the same phenotype will contribute to the probability of that phenotype to be transmitted, and thus influence is determined by the number of copiers that have already chosen a role-model with either phenotype.
Therefore, the probability of the $i$-th copier to copy phenotype $A$ is
\begin{equation}\label{eq:binary-model}
P_{i,A} = \frac{(N-X)\alpha'\beta(A) + K_{i,A}}{i-1 + (N-X)\alpha'\beta(A) + X\alpha'\beta(\hat{A})} = \frac{(N-X)\alpha'\beta(A) + K_{i,A}}{i-1 + (N-X)\alpha'\beta(A) + \alpha'X} % TODO should it be G_{i,A} rather than P_i,A ?
\end{equation}
where $X$ is the number of role-models with trait $\hat{A}$ and $K_{i,A}$ is the number of copiers that already chose $A$ when copier $i$ chooses a role-model and $\alpha'$ is the odd ratio of the bias weight $\alpha'= \frac{\alpha}{1-\alpha}$.

@@

\begin{corollary}
$\E[K_{Nj}] = \alpha_j \cdot \beta(A_j+e) / \overline{\alpha \cdot \beta(A+e)}$, where the averaging in the denominator is over the role-models index, $j$.
\end{corollary}

@@

% TODO what about a numerical validation for GBD?
% ANS - The results of this approximation are contained in the next DM approximation. In here we find the expected number of copiers for a single role-model, and in the DM we find that (and get the same results of course) and more. I can make the Numerical Validation a subsection and say it's for both.

@@

Examining variable values of the bias weight parameter $\alpha$, the test p-value was always 1, meaning that the null hypothesis cannot be rejected.
For high $\alpha$ values (above 0.5), very few simulations are needed to reach p-value of 1 (less than five simulations for $\alpha=0.9$, and less than 20 for $\alpha=0.7$)
For very low $\alpha$ values, which means very high influence weight, the number of simulations needed was 100.
When testing our observed distributions against different distributions, for example the uniform distribution (all role-models have exactly one copier), the p-value was 0, so that the null hypothesis was rejected, as required.
% TODO did you test other distributions?
% ANS - No, mainly because I didn't know what paremeters to choose, we should discuss

% TODO figure of observed vs expected distribution, and maybe also for other distributions such as uniform
% ANS - observed vs expected of what? Make a histogram of number of copiers per role-model? just the top role-models?

@@

Next, we examined the fixation probability and fixation time of an advantageous phenotype $\hat{A}$ when invading a population of phenotype $A$ and compared results from the full model and the DM approximation.
We find that the number of simulations needed to sufficiently approximate our model with the DM approximation is roughly $1,000$ (\Cref{fig:num_sims}).
Next, we examined the robustness of the DM approximation to relaxing the approximation assumptions.
First, we relaxed our assumption of no estimation error $e$.
Estimation error in the original model was drawn from a normal distribution, and added to the trait value before evaluation of the bias ($A_{ij} = A_j + e_i$).
When estimation error is applied, we sample $J_i$ for each copier $i$ from a normal distribution with varying scale (variance).
Even when the standard deviation is $0.1$, the fixation probability and time is similar~(\cref{fig:hetro_error}). 
We also relaxed our assumption of a uniform bias weight $\alpha$ (i.e., $\alpha_i=\alpha$). We allowed $\alpha$ to vary in the population, drawing $\alpha_j$ for each role-model $j$ from a normal distribution such that $\alpha_j \sim N(X,X)$. % TODO replace X and X with relevant symbols
We found again that results of the DM approximation are similar to those from stochastic simulations of the full model~(\cref{fig:hetro_alpha}).

\begin{figure}[h]
    \includegraphics[width=\linewidth]{../figures/binary/num_sims.pdf}
  \caption{
  \textbf{Number of simulations needed to get a good approximation.}
  The approximation (orange) fits simulation results (blue) well when using 1,000 simulations for both models. 
  Markers for average value across simulations, error bars for 95\% confidence interval.
  Here, population size, $N=1000$; bias weight, $\alpha=0.5$; phenotype values, $\hat{A}=1$, $A=0.7$; success bias value, $\beta(A)=0.956$.}	
  \label{fig:num_sims}
  % TODO
  % Kimura - black solid line
  % I removed J as you give beta(A) below, ok
  % it is not clear what "number of simulations needed" means.
  % what does "good enough" mean?
  % maybe y axis should be in relative error?
  % you should point to the equation for kimura.
\end{figure}

\begin{figure}
    \includegraphics[width=\linewidth]{../figures/binary/full_vs_dm_mutation.pdf}
  \caption{
  \textbf{Robustness of DM approximations to inclusion of estimation error.} Both the DM approximation (orange) and Kimura's equation (black line) fit the stochastic simulations (blue) well even with a high estimation error rate. Markers for average across simulations, error bars for 95\% confidence intervals.
  % and are condensed (+- 0.01 probability and +-40 generations) % what does that mean?
  5,000 simulations per data point; population size, $N=1000$; bias weight, $\alpha=0.1$; phenotype values, $\hat{A}=1$,$A=0.7$; $J\sim N(1,x^2)$ where $x$ is the estimation error scale in the x-axis. % not clear enough.
  }	
  % TODO
  % see comments in previous figure
  % maybe 50% CI looks better? the bars are too large to understand. maybe a wider y range? say from 0 to 0.02?
  \label{fig:hetro_error}
\end{figure}

@@

\begin{figure}[h]
%TODO remove the (a) and (b) from the latex and put them on the figure itself (top left corenr).
  \begin{center}
  \begin{subfigure}[a]{0.49\linewidth}
    \caption{}
    \includegraphics[width=\linewidth]{../figures/binary/fix_prob_var_alpha.pdf}
   \end{subfigure}
   \begin{subfigure}[a]{0.49\linewidth}
     \caption{}
    \includegraphics[width=\linewidth]{../figures/binary/fix_time_var_alpha.pdf}
   \end{subfigure}
  \end{center}
  \caption{Comparison of the DM approximation and the WF model for different values of the effective population size.
  Error bars show mean and 95\% confidence intervals.
  Effective population, $N_e=\alpha N + (1-\alpha)$.
  $5,000$ simulations per data point; Population size, $N=1,000$; $\hat{A}=1$, $A=0.7$,$J=1$,$1-\beta=s=0.044$.}
  % TODO
  % - better x and y label, as noted above
  % - in legend, replace "kimura" with "Approx."
  % - in legend, remove "=2s"
  % - change approximation color, it is too similar to WF simulations
  % - make the horizontal line black?
  % - caption should start with bold sentence summarizing the figure. 
  % - in caption add description of each parameter (as I did for N)
  \label{fig:var_alpha}
\end{figure}

@@

\begin{figure}[h]
  \begin{center}
  \begin{subfigure}[a]{0.49\linewidth}
  \caption{}
    \includegraphics[width=\linewidth]{../figures/changed_env/ch_env_var_k_div_l.pdf}
   \end{subfigure}
   \begin{subfigure}[a]{0.49\linewidth}
   \caption{}
    \includegraphics[width=\linewidth]{../figures/changed_env/ch_env_var_k_min_l.pdf}
   \end{subfigure}
  \end{center}
  %TODO add panel labels (a) and (b) on figures corner.
  % see previous figures todos
  % move legend to upper left? it is currently hiding something
  \caption{Model simulations compared with both the constant environment and the changing environment equations for different values of $k,l$, where $k$ is the number of generations where the bias is advantageous, and $l$ are the generations it is disadvantageous.
   When $k<l$ the approximation is good. When $k>l$, the approximation and the simulations are both very close to the constant environment approximation. 
 We ran $10,000$ simulations per data point, $N=1,000$, $\hat{A}=1$, $A=0.8$, $J=1$, $1-\beta=s=0.02$, $\alpha=0.1$.}
  \label{fig:ch_env_k_l}
\end{figure}

@@

\paragraph{Further work.} 
Our model can be expanded in various ways: observing the effects of different bias functions, % TODO unclear
including errors when estimating influence or costs when copying from an influential role-model, and inclusion of multiple, rather than a single, optimal phenotype.

>>>>

\begin{abstract}
Copying role-models can be an efficient method for acquiring knowledge. A common bias when choosing a role-model to copy is success bias: copying whoever appears more successful. This bias depends on the performance of the role-model alone, with no other factors. We propose an additional bias that may be prevalent in cultural transmission: influence bias, in which role-model choice is affected by the number of individuals that have already copied each potential role-model. We combine success and influence bias together to a ``prestige bias'' and analyze its effects on cultural-evolutionary dynamics using mathematical analysis and stochastic simulations. We find analytic approximations to our stochastic model, facilitating further mathematical analysis and reducing the computational complexity of simulations. We validate these approximations using simulations, and demonstrate their robustness to model assumptions.
We also find approximations to the fixation probability and the fixation time of an invading advantageous cultural trait, in both a constant and changing environment, which resemble Kimura's classical formulas.
These approximations show that success bias effectively plays the part of natural selection, whereas influence bias effectively reduces the population size.	
It also accelerates the evolutionary dynamics, as can be expected in a \textit{rich-getting-richer} process.
Our model may provide a better description of human cultural transmission, 
especially in the last years where social networks are very popular. % TODO to discuss this last sentence
Further work is needed to test if this model could predict various phenomena in human cultural evolution when extended with the effects of selection and innovation.
\end{abstract}
\pagebreak

\section*{Introduction}

@@

\paragraph{Success bias.}
\citet[Ch.8, p.247-249]{evolutionBook} describe a blended transmission algorithm by defining $F$ as a weighted average of the traits of all role-models, 
\begin{equation}\label{eq:boydF}
F_i(\vec{A}) = \sum_{j=1}^N G_{i,j}\cdot A_{i,j} \;, 
\end{equation}
where $G_{i,j}$ is the success bias of role-model $j$ in the eyes of copier $i$,
\begin{equation}\label{eq:boydG}
G_{i,j} = \frac{\beta(A_{i,j})}{\sum_{k=1}^{N} \beta(A_{i,k})} \;,
\end{equation}
$A_{i,j}$ is the absolute indicator trait value copier $i$ estimates for role-model $j$ with some error $e_i \sim N(0,\frac{1}{\eta^2})$,
\begin{equation}\label{eq:relativeIndicator}
A_{i,j} = A_j + e_i,
\end{equation}
and $\beta(\cdot)$ is the bias function that quantifies the success bias of a role-model,
\begin{equation}\label{eq:success_bias}
\beta(A_{i,j}) = b \cdot \exp{\Big(-\frac{(A_{i,j} - \hat{A})^2}{2J}\Big)},
\end{equation} 
with $\hat{A}$ as the arbitrary optimal indicator value, and $J$ and $b$ as parameters that control the bias strength.
Therefore, $G_{i,j}$ is a relative success score that copier $i$ assigns to role-model $j$, equivalent to relative fitness in evolutionary-genetic transmission models.

@@

\subsection*{Dichotomous trait}
To allow for mathematical analysis of the model, we introduce a simplified version where the indicator trait only has two phenotypes: the optimal phenotype $\hat{A}$ and the sub-optimal phenotype $A$. 
All role-models with the same phenotype will contribute to the probability of that phenotype to be transmitted, and thus influence is determined by the number of copiers that have already chosen a role-model with either phenotype.
Therefore, the probability of the $i$-th copier to copy phenotype $A$ is
\begin{equation}\label{eq:binary-model}
G_{i,A} = \frac{(N-X)\alpha'\beta(A) + K_{i,A}}{i-1 + (N-X)\alpha'\beta(A) + X\alpha'\beta(\hat{A})} = \frac{(N-X)\alpha'\beta(A) + K_{i,A}}{i-1 + (N-X)\alpha'\beta(A) + \alpha'X}
\end{equation}
where $X$ is the number of role-models with trait $\hat{A}$ and $K_{i,A}$ is the number of copiers that already chose $A$ when copier $i$ chooses a role-model and $\alpha'$ is the odd ratio of the bias weight $\alpha'= \frac{\alpha}{1-\alpha}$.

@@

\begin{corollary}\label{cor:GBD}
$\E[K_{Nj}] = \alpha_j \cdot \beta(A_j+e) / \overline{\alpha \cdot \beta(A+e)}$, where the averaging in the denominator is over the role-models index, $j$.
\end{corollary}

@@

\paragraph{Numerical validation.}
To validate our results we ran $1,000$ simulations of the full model, and compared the results with \cref{cor:GBD}.
We compare the distribution of number of copiers by plotting the histograms of both our simulations results and the expected values based on \cref{cor:GBD}.
We used the average of the $1,000$ simulations to eliminate drift errors, as can be seen in \cref{fig:GBD}.

\begin{figure}[h]
    \includegraphics[width=\linewidth]{../figures/chi_square_stats/GBD_validation.pdf}
  \caption{
  \textbf{Numerical validation for the GBD approximation.}
  The approximation (orange) fits simulation results (blue) well when using 1,000 simulations for both models.
  Here, population size, $N=2,000$; bias weight, $\alpha=0.1$; idea phenotype value, $\hat{A}=1$; role-model traits $\vec{A} \sim N(0,1)$; success bias value, $\beta(A)=0.956$.}	
  \label{fig:GBD}
\end{figure}

Although basic, \cref{fig:GBD} shows good fit of the GBD approximation.
This validation is initial, and the more extensive validations we do on the Dirichlet-Multinomial approximation, because it is what we will use in our analysts.

@@

Examining variable values of the bias weight parameter $\alpha$, the test p-value was always 1, meaning that the null hypothesis cannot be rejected.
For high $\alpha$ values (above 0.5), very few simulations are needed to reach p-value of 1 (less than five simulations for $\alpha=0.9$, and less than 20 for $\alpha=0.7$)
For very low $\alpha$ values, which means very high influence weight, the number of simulations needed was 100.
When testing our observed distributions against different distributions, for example the uniform distribution (all role-models have exactly one copier), the p-value was 0, so that the null hypothesis was rejected, as required.

@@

Next, we examined the fixation probability and fixation time of an advantageous phenotype $\hat{A}$ when invading a population of phenotype $A$ and compared results from the full model and the DM approximation.
We find that the number of simulations needed to sufficiently approximate our model with the DM approximation is roughly $1,000$ (\Cref{fig:num_sims}).
Next, we examined the robustness of the DM approximation to relaxing the approximation assumptions.
First, we relaxed our assumption of no estimation error $e$.
Estimation error in the original model was drawn from a normal distribution, and added to the trait value before evaluation of the bias ($A_{ij} = A_j + e_i$).
When estimation error is applied, we sample $J_i$ for each copier $i$ from a normal distribution with varying scale (variance).
Even when the standard deviation is $0.1$, the fixation probability and time is similar~(\cref{fig:hetro_error}). 
We also relaxed our assumption of a uniform bias weight $\alpha$ (i.e., $\alpha_i=\alpha$). We allowed $\alpha$ to vary in the population, drawing $\alpha_j$ for each role-model $j$ from a normal distribution such that $\alpha_j \sim N(0.5,x)$ where $x \in [10^{-7},10^{-1}]$. 
We found again that results of the DM approximation are similar to those from stochastic simulations of the full model~(\cref{fig:hetro_alpha}).

\begin{figure}[h]
    \includegraphics[width=\linewidth]{../figures/binary/num_sims.pdf}
  \caption{
  \textbf{Minimum number of simulations required to get a good approximation.}
  The approximation (orange) fits simulation results (blue) well when using 1,000 simulations for both models. The approximated value (black) is given by equation \cref{eq:kimura_p}.
  Markers for average value across simulations, error bars for 95\% confidence interval.
  Here, population size, $N=1000$; bias weight, $\alpha=0.5$; phenotype values, $\hat{A}=1$, $A=0.7$; success bias value, $\beta(A)=0.956$.}	
  \label{fig:num_sims}
\end{figure}

\begin{figure}
    \includegraphics[width=\linewidth]{../figures/binary/full_vs_dm_mutation.pdf}
  \caption{
  \textbf{Robustness of DM approximations to inclusion of estimation error.} Both the DM approximation (orange) and Kimura's equation (black line) fit the stochastic simulations (blue) well even with a high estimation error rate. Markers for average across simulations, error bars for 95\% confidence intervals.
  5,000 simulations per data point; population size, $N=1000$; bias weight, $\alpha=0.1$; phenotype values, $\hat{A}=1$,$A=0.7$; bias strength parameter $J\sim N(1,x^2)$ where $x \in [10^{-7},10^{-1}]$.
  }	
  % maybe 50% CI looks better? the bars are too large to understand. maybe a wider y range? say from 0 to 0.02?
  \label{fig:hetro_error}
\end{figure}

@@

\begin{figure}[h]
  \begin{center}
  \begin{subfigure}[a]{0.49\linewidth}
    \caption{}
    \includegraphics[width=\linewidth]{../figures/binary/fix_prob_var_alpha.pdf}
   \end{subfigure}
   \begin{subfigure}[a]{0.49\linewidth}
     \caption{}
    \includegraphics[width=\linewidth]{../figures/binary/fix_time_var_alpha.pdf}
   \end{subfigure}
  \end{center}
  \caption{Comparison of the DM approximation and the WF model for different values of the effective population size.
  Error bars show mean and 95\% confidence intervals.
  Effective population, $N_e=\alpha N + (1-\alpha)$.
  $5,000$ simulations per data point; Population size, $N=1,000$; $\hat{A}=1$, $A=0.7$,$J=1$,$1-\beta=s=0.044$.}
  % TODO
  % - better x and y label, as noted above
  % - in legend, replace "kimura" with "Approx."
  % - in legend, remove "=2s"
  % - change approximation color, it is too similar to WF simulations
  % - make the horizontal line black?
  % - caption should start with bold sentence summarizing the figure. 
  % - in caption add description of each parameter (as I did for N)
  \label{fig:var_alpha}
\end{figure}

@@

\begin{figure}[h]
  \begin{center}
  \begin{subfigure}[a]{0.49\linewidth}
  \caption{}
    \includegraphics[width=\linewidth]{../figures/changed_env/ch_env_var_k_div_l.pdf}
   \end{subfigure}
   \begin{subfigure}[a]{0.49\linewidth}
   \caption{}
    \includegraphics[width=\linewidth]{../figures/changed_env/ch_env_var_k_min_l.pdf}
   \end{subfigure}
  \end{center}
  %TODO add panel labels (a) and (b) on figures corner.
  % see previous figures todos
  \caption{Model simulations compared with both the constant environment and the changing environment equations for different values of $k,l$, where $k$ is the number of generations where the bias is advantageous, and $l$ are the generations it is disadvantageous.
   When $k<l$ the approximation is good. When $k>l$, the approximation and the simulations are both very close to the constant environment approximation. 
 We ran $10,000$ simulations per data point, $N=1,000$, $\hat{A}=1$, $A=0.8$, $J=1$, $1-\beta=s=0.02$, $\alpha=0.1$.}
  \label{fig:ch_env_k_l}
\end{figure}

@@

\paragraph{Further work.} %TODO revisit
Our model can be expanded in various ways: observing the effects of different bias functions, 
including errors when estimating influence or costs when copying from an influential role-model, and inclusion of multiple, rather than a single, optimal phenotype.
